Introducing an Ecological Trait-data Standard

### 2017-11-03 manuscript/manuscript.Rmd
Comments and additional references for the manuscript.

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07] and have been used to describe the role of species in ecosystems [@fontaine05] or their ability to persist under certain environmental conditions (@refs).

@@

Given the expected future challenges in dealing with climate change and managing land use, such global trait synthesis is an interdisciplinary/cross-topic field with great potential. 
<!-- Prospect: what will trait analysis acheive, if global data across taxa are available, -->

@@

Most prominently, plant traits from many different sources  have been collated into the quickly growing TRY database [@kattge11; e.g. including the extensive LEDA traitbase for the Northwest-European flora, @kleyer08].  Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] and or root measurements [Fine-root Ecology Database; @iversen17]. Traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17]. Trait-data for lower plants are less completely assessed. <!-- TRY collates trait data from other published databases with individually owned datasets, provides an extended access management, data query services and tracking of data usage in studies and secondary publications. The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage, which takes a huge effort to attract data submissions and map them into a common scheme. Repeating this exercise for other organism -->

@@

In the animal kingdom, specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems. For instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], or the Carabids.org web portal (http://www.carabids.org/). Besides these databases, an unknown plethora of single datasets has been published along with article publications over the decades [e.g., @gossner15, @gossner15c, @renner14; @birkhofer17; @vanderplas17  <!-- FDS: add some classics of trait data --> ]. In the past years, many datasets have been aggregated and harmonized into greater collections with particular organismal focus [e.g. @ricklefs17; @amphibio17; @iversen17; @myhrvold15; @kissling14; @jones09]. Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is lacking. It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits [@refs].

A global overview of existing trait data for all taxa and trait types is difficult to obtain. As an attempt to track  databases and unstandardized datasets from indipendent publications, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets with their their taxonomic focus, and extend of traits covered, their location on the internet and the terms of use. We invite data owners and users to register further datasets qualifying as trait data.

@@

<!-- 
(We conceptualize trait-datasets to  not include information about a species' interaction partners/associations of an organism, since they are -- to our understanding -- not within the scope of functional traits. However, such network data may be of interest to be combined with trait data. ) <!-- Malte: which can only be achieved if both types of datasets have been related to an ontology of species names. -->

@@

Some data-types may benefit from existing global identifiers for occurrence data, e.g. a GBIF URI or a museum collection code referencing the precise specimen at which the measuremetn was taken and that is publicly available.

@@

>>>>

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07] and have been used to describe the role of species in ecosystems [@fontaine05] or their ability to persist under certain environmental conditions [@Forsman2015] (@refs).

@@

Given the expected future challenges in dealing with climate change and managing land use, such global trait synthesis is an interdisciplinary/cross-topic field with great potential. 
<!-- Prospect: what will trait analysis acheive, if global data across taxa are available, -->
<!-- Nadja: Some thoughts/snippets based on other publications: " a relevant amount of data is available to quantify and understand trait variation beyond aggregation on species level. The analysis presented here is only a first step to disentangle within- and between-species variability. [...] A more empirically based representation of functional diversity is expected to contribute to an improved prediction of biome boundary shifts in a changing environment. [...] Empirical inference of this kind is expected to stimulate and strengthen hypothesis-driven research (Golub, 2010; Weinberg, 2010) towards a unified ecological assessment of plant traits and their role for the functioning of the terrestrial biosphere." [@kattge11] / "Our traitbased approach to modelling competition makes generalization possible across the forest ecosystems of the world and their highly diverse species composition. [...] A challenge for the future is to move beyond growth to analyse all key demographic rates and life-history stages, and analyse how traits influence competitive outcomes and stable coexistence at the population level."[@Kunstler2016] / "Traits include those that in combination allow a first-order assessment of species' similarity in "ecological niches" or the overlap in their functional role. The standardized interpretation and adherence to source information and clear documentation aims to provide a replicable and transparent basis for research. [...] The trait variables in the dataset should support hypothesis testing or applied inference in ecology, conservation and evolution." [@Wilman2014] -->

@@

Most prominently, plant traits from many different sources  have been collated into the quickly growing TRY database [@kattge11; e.g. including the extensive LEDA traitbase for the Northwest-European flora, @kleyer08].  Other more specialised databases for plant traits exist, for instance flower traits [BiolFlor for German wild flowers, @klotz02] and root measurements [Fine-root Ecology Database; @iversen17]. Traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17]. Trait-data for lower plants are less completely assessed. <!-- TRY collates trait data from other published databases with individually owned datasets, provides an extended access management, data query services and tracking of data usage in studies and secondary publications. The great achievement of TRY is to harmonize these data into a unified scheme and provide a single platform of global coverage, which takes a huge effort to attract data submissions and map them into a common scheme. Repeating this exercise for other organism --> 
<!-- Nadja: This is of course the major achievement of TRY, but I think we don't have to mention it because this not our immediate goal with the trait standard. The important bit for us is that we provide tools to harmonize datasets. -->

@@

In the animal kingdom, specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems. For instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/). Besides these databases, an unknown plethora of single datasets has been published along with article publications [e.g., @gossner15, @gossner15c, @renner14; @birkhofer17; @vanderplas17  <!-- FDS: add some classics of trait data --> @Sokol2011, @Price2013 <!-- Nadja: oldest relevant pub. in Dryad-->]. In the past years, many datasets have been aggregated and harmonized into greater collections with particular organismal focus [e.g. @ricklefs17; @amphibio17; @iversen17; @myhrvold15; @kissling14; @jones09]. Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is lacking. It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits [@refs].

A global overview of existing trait data for all taxa and trait types is difficult to obtain. As an attempt to track  databases and unstandardized datasets from indipendent publications, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets with their their taxonomic focus and extend of traits covered, their location on the internet and the terms of use. We invite data owners and users to register further datasets qualifying as trait data.

@@

<!-- 
(We conceptualize trait-datasets to  not include information about a species' interaction partners/associations of an organism, since they are -- to our understanding -- not within the scope of functional traits. However, such network data may be of interest to be combined with trait data. ) <!-- Malte: which can only be achieved if both types of datasets have been related to an ontology of species names. --> <!--Nadja: I don't think that we need to mention this. If you think that this is important, it should be moved more to the beginning, e.g. within the section "Common terms for trait data"-->

@@

Some data-types may benefit from existing global identifiers for occurrence data, e.g. a GBIF URI or a museum collection code referencing the precise specimen at which the measurement was taken and that is publicly available.

@@

<!-- Nadja: can those last two paragraphs be merged with the second paragraph? I feel like we are repeating ourselves here. -->
